---
title: "RuhiApp"
author: "Tyler Marino"
date: "2024-03-14"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("maps")       #Must install maps

library(maps)                  #Maps
library(leaflet)               #Plotting on maps
library(dplyr)                 #dataframes dealings 
library(shiny)                 #app making
library(readxl)
library(stringr)

```

This R Markdown document is made interactive using Shiny. Unlike the more traditional workflow of creating static reports, you can now create documents that allow your readers to change the assumptions underlying your analysis and see the results immediately. 

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Functions and Fundamental needs
```{r}

#Create function to make degree minutes into degrees INPUT(degree minutes) OUTPUT(degree)
  #Ensure that if the degree is negative that you turn the minute negative aswell.
redefine_degree <- function(degree_minute_decimal) {
  
  # Splitting the string into degree, (minute and decimal minute)
  split_components <- strsplit(degree_minute_decimal, "[° ]")[[1]]

  #Extracting degree
  degree <- as.numeric(split_components[1])
  if (degree < 0){
    minute <- as.numeric(split_components[3])
    minute <- -1*minute/60
  }
  else {
    minute <- as.numeric(split_components[3])
    minute <- minute/60
  }

  #Output the results
  degree <- degree + minute
  return (degree)
}



#Function to make the df contain the new degrees. INPUT(df), OUTPUT(df with lat, long)
change_df_degrees <- function(df){
  df$lat <- sapply(df$Latitude, redefine_degree)
  options(digits = 20)
  df$long <- sapply(df$Longitude, redefine_degree)
  options(digits = 20)
  return (df)
}



#DICTIONARY of colors for each section of the Delta System
category_colors <- c("South Bay" = "darkred", 
                     "Central Bay" = "blue", 
                     "San Pablo Bay" = "green", 
                     "West Suisun Bay" = "orange", 
                     "East Suisun Bay" = "purple", 
                     "Confluence" = "white", 
                     "Lower Sacramento River" = "yellow", 
                     "Lower San Joaquin River" = "black")



#Fucntion df to have associated Icon colors wrt to Area of Interest INPUT(df) OUTPUT(df wiht iconColors column)

addColortoDF <- function(df) {
  df$iconColor <- unlist(category_colors[match(df$`Area of Interest`, names(category_colors))])
  return (df)
}

#Function to create a custom icon based on color of area of interest INPUT(color) OUTPUT(Marker Icon)
custom_icon <- function(color) {
  awesomeIcons(
      icon = 'star',
      markerColor = color,
      iconColor = 'grey',
      library = 'ion'
      )
}

#Function to get color for ICON INPUT(df), OUTPUT(color)
getColor <- function(df) {
  sapply(df$Region, function(Region) {
  if(Region == "South Bay") {
    "darkred"
  } else if(Region == "Central Bay") {
    "blue"
  } else if(Region == "San Pablo Bay"){
    "green"
  } else if(Region == "West Suisun Bay") {
    "orange"
  } else if(Region == "East Suisun Bay"){
    "purple"
  } else if(Region == "Confluence") {
    "white"
  } else if(Region == "Lower Sacramento River") {
    "yellow"
  } else {
    "black"
  }
    })
}


```


##LOAD DATA
```{r}
stations_df1 <- read_excel("C:\\Users\\Tyler Marino\\OneDrive\\Desktop\\RuhiProject\\StationPoints\\StationPoints1.xlsx")

stations_df_test <- read_excel("C:\\Users\\Tyler Marino\\OneDrive\\Desktop\\RuhiProject\\StationPoints\\StationPoints1.xlsx")
```

##MANIPULATE DATA
```{r}



#Add lat and long to df
stations_df1 <- change_df_degrees(stations_df1)

#Add color to df, now I have test df
stations_df1 <- addColortoDF(stations_df1)

##REMOVE NAs from the data
stations_df1 <- na.omit(stations_df1)

#change area of interest column name to 'Region'
stations_df1 <- stations_df1 %>% 
       dplyr::rename("Region" = "Area of Interest")
```



## Talk to Me @Shiny App
```{r}
#Create the icons
iconss <- lapply(stations_df1$iconColor, custom_icon)
icons <- awesomeIcons(
  icon = 'ios-close',
  iconColor = 'grey',
  library = 'ion',
  markerColor = getColor(stations_df1)
)

#user interface with inputs and outputs
  #Title
  #Sidebar
    #For selecting Region: dropdown
      #In progress: MultiInput for regions instead of dropdwon
  # Main Panel (Map)
ui <- fluidPage(
  titlePanel("Sacramento/San Joaquin Delta System Stations"),
  
  
  sidebarLayout(
    sidebarPanel(
      selectInput("RegionColor", "Region Of Bay:", 
                  choices = stations_df1$Region, selected = "Central Bay"),

    ),
  # multiInput(
  #   inputId = "RegionInput", label = "Regions :",
  #   choices = stations_df1$Region,
  #   selected = NULL, width = "400px",
  # ),
    
    
    mainPanel(leafletOutput("BayDelta"))
  )

)



#Server, the backend where the magic is made

  #First select only regions of DF we want. Need to       make  Reactive
  #We must call to render leaflet map
  #Then we construct the map
    #Tiles
    #Setting start location of map
    #Add Markers (FAULTY)
server <- function(input, output, session) {
  df_to_display <- stations_df1
  # df_to_display <- reactive({
  #   stations_df1[stations_df1$`Area of Interest` == input$Region,]
  # })
  
  output$BayDelta <- renderLeaflet({                
    
    leaflet(df_to_display) %>%
      
      addTiles() %>%
      
      setView(lng = mean(df_to_display$long), lat = mean(df_to_display$lat), zoom = 8) %>%
      
      addAwesomeMarkers(
        lng = ~long,
        lat = ~lat, 
        icon = icons, 
        label = df_to_display$`Station Number`,
        popup = ~Description
      )
  })
}


#Call the shiny app
shinyApp(ui, server)
```

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change.  This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot.


```{r eruptions, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r tabsets, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package = "shiny"),
  options = list(
    width = "100%", height = 550
  )
)
```

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.



